---
title: "EDA- World GDP vs Labour Markets"
author: "Ntandoyenkosi Matshisela"
date: "18/10/2021"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r packages}

# Libraries


suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(gapminder)))
suppressMessages(suppressWarnings(library(cowplot)))
suppressMessages(suppressWarnings(library(DT)))
suppressMessages(suppressWarnings(library(downloadthis)))
suppressMessages(suppressWarnings(library(ggstatsplot)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(WDI)))
```

## World Bank Indicators

The World Bank has a data repository that has energy, education, financial etc
indicators. For this analysis, 4 indicators will be extracted and an exploratory
data analysis will be done. The four indicators of interest are Population,
Unemployment, Labour (Total) and lastly the GDP in $US. To download an R package
known as WDI will be used.

To help in investigating the data, a simple feature engineering is done. The 
percentage of the Working class (Labour force divided by Population) is 
determined. 

```{r }

# Accessing World Bank Indicators
data <- WDI(
  country = "all",
  indicator = c("Population" = "SP.POP.TOTL",  # Population
                "Unemployment" = "SL.UEM.TOTL.ZS", # Unemployment, 
                "Labour" ="SL.TLF.TOTL.IN", # Labour Force
                "GDP in US$" = "NY.GDP.MKTP.CD"), # GDP in US$
  start = 1960,
  end = 2020,
  extra = TRUE,
  cache = NULL,
  latest = NULL,
  language = "en"
)


data1 <- data %>%
  filter(region != "Aggregates" & year >= 1990) %>%
  select(country, region, year, `GDP in US$`, Population, Unemployment, Labour) %>%
  mutate(Working_class = Labour/Population) # Feature Engineering

data1 %>%
  datatable(extensions = 'Buttons',
            options = list(dom = 'lfrtipB',
                           buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                           lengthMenu = list(c(10,25,50,-1),
                                             c(10,25,50,"All"))))

```
# Labour Markets

As expected, countries with high populations have labour forces. China and India
have recorded high population statistics over the last 30 years, although China 
rate of increase has reduced drastically since 2005. China GDP is fast 
approaching that of the US. The outlier points in GDP and Labour makes the data
too clustered for one to carefully notice the movement of other countries. There
are a lot of inferences that can be hypothesized or suspected, such as the 
relationship between the 2 variables. 

```{r, fig.height= 10, fig.width= 10}

# Labour Plot
x <- ggplot(data1, aes(`GDP in US$`, Labour, color = region)) +
  geom_point(aes(size = Labour, frame = year, ids = country)) +
  scale_x_log10() + theme_bw() + labs(title = "GDP vs Labour")

ggplotly(x)

```

# Working Class

I would like to investigate the percentage of the population that is independent.
That is, the group of individuals that are working in each country. Well some 
jobs pay far less than the poverty datum line, making one dependent to a parent
or a spouse etc. Countries such as Qatar, United Arab Emirates, Singapore and 
New Zealand have recorded high levels of migration of working class individuals
in the last 30 years.

```{r, fig.height= 10, fig.width= 10}

# Working Class % 
z <- ggplot(data1, aes(`GDP in US$`, Working_class, color = region)) +
  geom_point(aes(size = Working_class, frame = year, ids = country)) +
  scale_x_log10() + theme_bw() + labs(title = "GDP vs Working Class")

ggplotly(z)

```

# Unemployment

Around 2008 almost all countries recorded high levels of unemployment after the
2008 market crash. It can be observed also that although South Africa has one
of the highest GDP in Africa it has the highest percentages of unemplyment. 
Some scholars attribute the crime rate to inequalities bedevilling the country. 
China with the highest population in the world has one of the lowest unempoyment
rates, more investigations can be explored here to translate lessons to other 
nations.

```{r, fig.height= 10, fig.width= 10}
# Unemployment
r <- ggplot(data1, aes(`GDP in US$`, Unemployment, color = region)) +
  geom_point(aes(size = Unemployment, frame = year, ids = country)) +
  scale_x_log10() + theme_bw() + labs(title = "GDP vs Unemployment")

ggplotly(r)

```

Some scolars argue that GDP solves the unemployment crisis. To investigate this,
the following analysis is conducted.

```{r, fig.height= 10, fig.width= 10}

set.seed(123) # reproducibility

data2 <- data1 %>%
  filter(year >= 1990)


# plot
grouped_ggscatterstats(
  # arguments relevant for ggstatsplot::ggscatterstats
  data = data2,
  x = `GDP in US$`,
  y = Unemployment,
  grouping.var = region,
  label.var = country,
  label.expression = `GDP in US$` > 100000000000 & Unemployment > 27,
  type = "r",
  ggtheme = ggthemes::theme_tufte(),
  # arguments relevant for ggstatsplot::combine_plots
  annotation.args = list(
    title = "Relationship between GDP and Unemployment",
    caption = "Source: https://data.worldbank.org/indicator"
  ),
  plotgrid.args = list(nrow = 4, ncol = 2)
)


```

It can be seen that there is no statistical evidence that GDP has an effect
on unemployment in some regions in the world.

This analysis exposed that the world has high labour markets and in cases with 
lower or high paying, migration is recorded. The analysis further exposed the
challenges of unemployment and that GDP has no effect on employment in some 
countries, while in some, it has indeed shown improvement in this indicator.